\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-25 document/XSGRB_sample.tex
Comments from Giorgos and Jens + notes on individual bursts

@@

\title{XSGRB I - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

\abstract{In this work we present the spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	\% of these, covering a redshift range from 0.059 to 7.84. Based on a set of observational selection criteria that are unbiased with regards to intrinsic properties of the
	GRBs, the follow-up effort has been guided by the \textit{Swift}-telescope to
	produce a homogeneous sample of afterglows for 86 spectra. We here provide a
	public release of all the reduced spectra, including continuum estimates and
	telluric corrections. For completeness, we also provide reductions for the
	late-time observations of the underlying host galaxies. We provide an assessment
	of the degree of completeness with respect to the parent GRB population, in
	terms of the X-ray properties of the bursts in the sample and find that the
	sample presented here is consistent with being drawn from the \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the neutral hydrogen column density, increasing the total number of published HI
	column density measurements by $\sim$ 50 per cent. We constrain the fraction of
	dark bursts to be less than 28 per cent based on the detection of an optical
	afterglow and tentatively confirm previous results that larger optical darkness
	is correlated with increased X-ray absorption. This dataset provides a unique
	possibility to study the ISM across cosmic time. }

@@

$\gamma$-ray bursts (GRBs) are interesting as they provide the opportunity to
derive new constraints on a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Lyons2010, Molinari2007, HjorthBloom2012, Wijers1998,
	Prochaska2009, Savaglio2006, Ghirlanda2007}. In that sense GRB research is very
multidisciplinary.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based follow-up
observations of the afterglows and of the host galaxies of the GRBs. This work
has involved a large community of researchers both inside and outside the
official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12 degrees.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No closeby bright stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Our ability to observe GRB afterglows is strongly dependent on the timing and
the precision of the target positions delivered by the triggering facilities. By
selecting only bursts that have been triggered on board the \textit{Swift} space
telescope \citep{Gehrels2004}, based on the Burst Alert Telescope (BAT)
\citep{Barthelmy2005}, we  start out with a homogeneously selected sample where
burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006, Coward2013a}, attempts at inferring properties
of the underlying GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016}. Restricting
the follow-up effort to bursts detected by \textit{Swift}, we therefore ensure
that the limitations of the parent sample are well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, a secure host identification is in general impossible
based on BAT alone. We therefore additionally require an X-ray position from the
X-Ray Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} within 12 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high
\citep{Burrows2007}, this cut should not alter the parent sample significantly.

@@

Under rare circumstances, the use of the rapid response mode (RRM) has been
possible. In case of a promptly visible GRB, a robotic trigger was sent to
the telescope if at the moment of the GCN notice the GRB is observable and
fulfills the following criteria.

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope
	\item GRB promptly visible
	\item Error radius < 60\arcsec.
	\item No limit on declination. 
	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky > 22$^\circ$ (both now and 15 min after the trigger).
	\item Sun elevation < -12$^\circ$ (both now and 15 min after the trigger).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The flags KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true).
\end{enumerate}

@@

There 8 GRBs in the sample that were being observed with the RRM mode. One of
the RRM triggers is outside the statistical sample and two of the RRM trigger
are on short GRBs. The fastest response between \swift~GRB trigger and the
beginning of spectroscopic integration is for GRB~160410A, for which the delay
was only 8.4 minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red circles are added around GRBs which have been
	followed up by X-shooter since the commissioning of the instrument. Blue points
	with red circles represent GRBs that enter our sample and have X-shooter
	spectroscopy. The different samples are compared in Sect. \ref{results}. The
	background shows the dust maps presented in \citet{Schlegel1998} where the
	sample criteria cut with $A_V \lesssim 0.5$ mag is removed. The colorbar below
	indicates the value of $A_V$ on the plot. The dotted lines indicate intervals
	of 30$^\circ$ in longitude and latitude.} \label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of 8 years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. We
have included a few additional bursts, from the programmes 092.D-0056(A) (PI:
Rau), 092.D-0633(E) (PI: Greiner), and 095.B-0811(B) (PI: Levan). The total
collection of spectra represents \textit{all} GRBs that have been followed up by
X-shooter up to \termdate.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high Galactic extinction cutoff (marked in
white), the GRB positions have an apparent isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

A subset of the spectra presented here, primarily host observations, have
already been published in \citet{Kruhler2015}. We include independent
reductions of them here for completeness.

@@

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that the Moffat function
\citep{Moffat1969} adequately describes an imaging PSF due to atmospheric
turbulence, but because of aberrations in the dispersion elements and the
rectification process, the PSF we are trying to model is different from this
profile. To allow for flexibility in the model, we have chosen the Voigt
function as a model for the spectral PSF and we describe how this is evaluated
in App. \ref{voigt}. Since additionally, the host galaxy could also have a
contribution to the image profile, this choice allows for the required freedom if
additional flux is in the wings of the profile.

@@

For all earth-based telescopes, the light first has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritized list of methods here, depending on the
availability of the different methods. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the optimal correction.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have overplot the most
	prominent lines seen in GRB afterglows from \citet{Christensen2011a}.
	Additionally, three intervening systems are seen in this sightline This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is not terribly accurate (for one, the spatial profile of the
trace is not a perfect Gaussian), it nevertheless gives a sufficiently accurate
estimate for the analysis of these poor signal to noise science spectra.

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRBs after \startdate, that
have been observed with X-shooter, while only a subset of these constitutes our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.
curiosity properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a burst outside the statistical sample is the
bright \textit{INTEGRAL} burst GRB~161023A, that turned to be a rich spectrum with at
least 11 intervening absorption systems.

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption line
imprinted on the GRB afterglow continuum. Because the afterglow rapidly fades
(typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations in
our sample is for the short, $z = 1.717$, GRB~160410A for which spectroscopic
integration was initialed only 8.4 minutes after the BAT trigger. To illustrate
importance of the follow-up delay for the redshift completeness, we plot the
redshift completeness as a function of delay time in Fig. \ref{fig:timing} for
all the bursts we have followed up, including the ones outside the statistical
sample. Visible is the decreasing fraction of GRBs with a redshift determination
for increasing delays. The overall redshift completeness for bursts that we have
followed up is 92 per cent. This degree of completeness in the followed bursts,
illustrates the efficiency of VLT/X-shooter in redshift determination. Not shown
in the figure are an additional 12 bursts that have redshift determinations
based on late-time host observations with delay times longer than $\sim$10 days.

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/timing.pdf}} \caption{Afterglow
	magnitude at the start of observation as a function of follow-up delay. The
	points have been colored based on the redshift of the corresponding burst. Red
	crosses indicate GRBs without a measured redshift and arrows indicate bursts
	for which the afterglow was not detected in the acquisition image. In red is
	shown the redshift completeness as a function of follow-up delay.}
\label{fig:timing}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample completeness} \label{completeness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
bursts have triggered \textit{Swift} in the period since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 60 per cent. We return to the
question of redshift completeness in Sect. \ref{redshift}. From this sample,
90 GRBs have been spectroscopically followed up with X-shooter. In order to
assess whether the subset of bursts followed up are representative of the
underlying GRB parent population, we compare intrinsic properties of GRBs in our
sample to GRBs in the full sample followed up by \textit{Swift}. We show the
comparison between the BAT (15-150 keV) fluence, the XRT flux (0.3 - 10 keV) at
11 hours, and the intrinsic X-ray derived equivalent hydrogen column density,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the bursts. We return to this last
point in Sect. \ref{darkness}.

\begin{figure*}
	\centerline{\includegraphics[width=18cm]{figures/completeness_BAT.pdf}}
\caption{Comparison between burst properties of all bursts observed with
	\textit{Swift} (blue), the ones that fullfill the sample criteria specified in
	Sect. \ref{samplecrit} (green), and the subset that has been observed as part of
	the statistical sample (red). The left plot shows the fluence in 15-150
	keV as observed by BAT. The middle panel shows the 0.3 - 10 keV flux, 11 hours
	after the bursts as measured by XRT. In the right most panel, we show
	$N_{\mathrm{HI,X}}$ based on the XRT spectrum \citep{Evans2009}.}
\label{fig:swift_complete}
\end{figure*}

@@

For each of the samples, we calculate the median, 14th, and 86th percentiles of
each of the distributions, which can be used as point estimates for the
population distribution. These are provided in Table.
\ref{tab:sample_properties}. It can be seen from the values that the three
samples have very similar distributions in terms of the point estimates chosen.
This suggests that our selection criteria are unbiased compared to the
\textit{Swift}-sample and that additionally, the follow-up effort conserves the
distributions of the intrinsic GRB properties (except perhaps for $N_{\mathrm{HI,X}}$, see Sect. \ref{darkness}).

@@

\subsection{GRB~161023A (z = 2.710)}

>>>>

\usepackage{float}

@@

\title{XS-GRB I - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

@@

\abstract{In this work we present the spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that are unbiased with regards to intrinsic
	properties of the GRBs, the follow-up effort has been guided by the
	\textit{Swift}-telescope to produce a homogeneous sample of spectra for 91 GRBs.
	We here provide a public release of all the reduced spectra, including continuum
	estimates and telluric corrections. For completeness, we also provide reductions
	for the late-time observations of the underlying host galaxies. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is representative of the full \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the neutral hydrogen column density, increasing the total number of published
	HI column density measurements by $\sim$ 50 per cent. %	We constrain the
	% fraction
	%	of dark bursts to be less than 28 per cent based on the detection of an
	% optical
	%	afterglow and tentatively confirm previous results that larger optical
	% darkness
	%	is correlated with increased X-ray absorption.
	This dataset provides a unique possibility to study the ISM across cosmic time,
	from the local progenitor surroundings to the galaxy wide conditions}

@@

$\gamma$-ray bursts (GRBs) provide the opportunity to derive new constraints on
a very wide range of topics in astrophysics. Examples range from small scale
phenomena relating to magnetars, properties of highly relativistic jets,
hyper/supernova explosions, the interstellar medium, dust extinction curves,
starburst galaxies, chemical and molecular abundances, escape of ionizing
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology \citep[e.g.,][]{Lyons2010,
	Molinari2007, HjorthBloom2012, Wijers1998, Prochaska2009, Savaglio2006,
	Ghirlanda2007, Vergani2009, Christensen2017}.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. With the two instruments, the
Burst Alert Telescope (BAT) \citep{Barthelmy2005} and the X-Ray Telescope (XRT)
\citep{Burrows2005}, \swift~is an ideal starting point for GRB investigations. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been greatly facilitated by the open data access policy of the
{\it Swift} mission.

@@

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12 degrees.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright, nearby stars within $ 1.8 + 0.4 \times exp[(20 - R)/2.05$] arcsec, where R is the USNO magnitude of the star)
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Our ability to observe GRB afterglows is strongly dependent on the timing and
the precision of the target positions delivered by the triggering facilities. By
selecting only bursts that have been triggered on board the \textit{Swift} space
telescope \citep{Gehrels2004}, based on the BAT, we  start out with a homogeneously selected sample where
burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006, Coward2013a}, attempts at inferring properties
of the underlying GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016}. Restricting
the follow-up effort to bursts detected by \textit{Swift}, we therefore ensure
that the limitations of the parent sample are well studied.

Because the localization accuracy of BAT is 1 -- 4 arcminutes
\citep{Barthelmy2005}, a secure host association is in general very hard based
on BAT alone. We therefore additionally require an X-ray position from XRT to be
distributed to the GCN network \citep{Barthelmy2000} within 12 hours and to
account for observing constraints on \textit{Swift} that XRT began observations
within 10 minutes. The additional timing requirement of the XRT follow-up means
that all bursts in our sample have detected X-ray afterglows. Because the XRT
completeness is very high \citep{Burrows2007}, this cut should not alter the
parent sample significantly.

@@

Under rare circumstances, the use of the ESO rapid response mode (RRM) has been
possible. In case of a promptly visible GRB, a robotic trigger was sent to the
telescope if at the moment of the GCN notice the GRB is observable and fulfills
the following criteria.

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope
	\item GRB promptly visible
	\item Error radius < 60\arcsec.
	\item No limit on declination. 
	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky > 22$^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation < -12$^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The flags \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true).
\end{enumerate}

@@

There are 8 GRBs in the sample that have beem being observed with the RRM mode.
One of the RRM triggers is outside the statistical sample and two of the RRM
trigger are on short GRBs. The fastest response between \swift~GRB trigger and
the beginning of spectroscopic integration is for GRB~160410A, for which the
delay was only 8.4 minutes.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red outline is added around GRBs which have been followed
	up by X-shooter in this work. Blue symbols with red outlines represent GRBs
	that enter our sample and have X-shooter spectroscopy. The different samples
	are compared in Sect. \ref{results}. The background shows the dust maps
	presented in \citet{Schlegel1998} where the sample criteria cut with $A_V
	\lesssim 0.5$ mag is replaced with a white color. The colorbar below indicates
	the value of $A_V$ on the plot. The dotted lines indicate intervals of
	30$^\circ$ in longitude and latitude.} \label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of 8 years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260 (PI:
Fynbo). We have included a few additional bursts, from the programmes
084.D-0265(A) (PI: Benetti), 091.A-0877(A) (PI: Schady), 092.D-0056(A) (PI:
Rau), 092.D-0633(E) (PI: Greiner), and 095.B-0811(B) (PI: Levan). The total
collection of spectra represents \textit{all} GRB afterglows that have been
followed up by X-shooter up to \termdate.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high Galactic extinction cutoff (marked in
white), the GRB positions have an isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

30 percent of the spectra presented here, primarily host observations, have
already been published in \citet{Kruhler2015}. We include independent
reductions of them here for completeness.

@@

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that a Moffat function
\citep{Moffat1969} with an index of $\beta = 4.765$ adequately describes an
imaging PSF due to atmospheric turbulence, but because of aberrations in the
dispersion elements and the rectification process, the PSF we are trying to
model is different from this profile. To allow for flexibility in the model, we
have chosen the Voigt function as a model for the spectral PSF and we describe
how this is evaluated in App. \ref{voigt}. Since additionally, the host galaxy
could also have a contribution to the image profile, this choice allows for the
required freedom if additional flux is in the wings of the profile.

@@

For all earth-based telescopes, the light first has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritized list of methods here, depending on the
availability of the different methods. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the best possible correction.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have overplot the most
	prominent lines seen in GRB afterglows from \citet{Christensen2011a}.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is perfect (the spatial profile of the trace is not a perfect
Gaussian), it nevertheless gives a sufficiently accurate estimate for the
analysis of these poor signal to noise science spectra.

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRBs after \startdate, that
have been observed with X-shooter, while only a subset of these constitutes our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.
curiosity properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a spectacular burst outside the
statistical sample is the bright \textit{INTEGRAL} burst GRB~161023A, that
contains at least 11 intervening absorption systems, see \ref{GRB161023A}.

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption line
imprinted on the GRB afterglow continuum. Because the afterglow rapidly fades
(typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations in
our sample is for the short, $z = 1.717$, GRB~160410A for which spectroscopic
integration was initiated only 8.4 minutes after the BAT trigger. To illustrate
the importance of the follow-up delay for the redshift completeness, we plot the
redshift completeness as a function of delay time in Fig. \ref{fig:timing} for
all the bursts we have followed up, including the ones outside the statistical
sample. As can be seen from the figure, the fraction of GRBs with a redshift
determination decreases with follow-up delay. The overall redshift completeness
for bursts that we have followed up is 92 per cent. This degree of completeness
in the followed bursts, illustrates the efficiency of VLT/X-shooter in redshift
determination. Not shown in the figure are an additional 12 bursts that have
redshift determinations based on late-time host observations with delay times
longer than $\sim$10 days.

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/timing.pdf}} \caption{Afterglow
	magnitude at the start of observation as a function of follow-up delay. The
	points have been colored based on the redshift of the corresponding burst. Red
	crosses indicate GRBs without a measured redshift and arrows indicate bursts
	for which the afterglow was not detected in the acquisition image. In red is
	shown the redshift completeness as a function of follow-up delay.}
\label{fig:timing}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample completeness} \label{completeness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
bursts have triggered \textit{Swift} in the period since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 61 per cent. We return to the
question of redshift completeness in Sect. \ref{redshift}. From this sample, 91
GRBs have been spectroscopically followed up with X-shooter. In order to assess
whether the subset of bursts followed up are representative of the underlying
GRB parent population, we compare intrinsic properties of GRBs in our sample to
GRBs in the full sample followed up by \textit{Swift}. We show the comparison
between the BAT (15 -- 150 keV) fluence, the XRT flux (0.3 -- 10 keV) at 11
hours, and the intrinsic X-ray derived equivalent hydrogen column density,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the full \swift~sample. We return to
the last point in Sect. \ref{darkness}.

\begin{figure*}
	\centerline{\includegraphics[width=18cm]{figures/completeness_BAT.pdf}}
\caption{Comparison between burst properties of all bursts observed with
	\textit{Swift} (blue), the statistical sample that fullfill the criteria
	specified in Sect. \ref{samplecrit} (green), and the subset that has been
	observed as part of the statistical sample (red). The left plot shows the
	fluence in 15-150 keV as observed by BAT. The middle panel shows the 0.3 - 10
	keV flux, 11 hours after the bursts as measured by XRT. In the right most
	panel, we show $N_{\mathrm{HI,X}}$ based on the XRT spectrum
	\citep{Evans2009}.} \label{fig:swift_complete}
\end{figure*}

@@

For each of the samples, we calculate the median, 14th, and 86th percentiles of
each of the distributions, which can be used as point estimates for the
population distribution. These are provided in Table
\ref{tab:sample_properties}. It can be seen from the values that the three
samples have very similar distributions in terms of the point estimates chosen.
This suggests that our selection criteria are unbiased compared to the
\textit{Swift}-sample and that additionally, the follow-up effort conserves the
distributions of the intrinsic GRB properties (except perhaps for
$N_{\mathrm{HI,X}}$, see Sect. \ref{darkness}).

@@

\subsection{GRB~161023A (z = 2.710)}	 \label{GRB161023A}
